---
title: "Data Clean"
format: html
editor: visual
---

## Setup

Loading required packages.

```{r}
#| message: false
library("tidyverse")
library("dplyr")
```

Loading the data.

```{r}
#| message: false
count_data <- read_tsv("../data/01_dat_load_count.tsv")
meta_data <- read_tsv("../data/01_dat_load_meta.tsv")
```

## Data Wrangling I

In this section, the data variables are cleaned up for count data and meta data seperately.

### Gene Expression Counts

We see that the data do not only consist of gene counts but is also interrupted by p values.

```{r}
count_data |> 
  sample_n(10)
```

The detected p values are removed such that only the counts are left.

```{r}
count_data <- count_data |> 
  dplyr::select(-matches(match = "Detection.Pval"))

count_data |> 
  sample_n(10)
```

Column names are simplified as the date is saved in the meta data.

```{r}
count_data |> 
  colnames() = str_extract(string = colnames(count_data),
                           pattern = "^[A-Z0-9]+\\.?[0-9]*")

count_data |> 
  sample_n(10)
```

### Meta Data

We see what meta data is available for the samples.

```{r}
colnames(meta_data)
```

We will be interested in keeping title, geo_accession, and description.1. The last mentioned column is the identifier fom each sample that is found in the data for expression counts. Furthermore, we want information about gender and disease state described by characteristics_ch1 and characteristics_ch1.2. This information is also found in the variables gender:ch1 and disease state:ch1 in a nicer format. We chose to work with the first mentioned columns for educational purposes.

```{r}
meta_data <- meta_data |>
  dplyr::select(title,
                geo_accession,
                characteristics_ch1,
                characteristics_ch1.2,
                description.1) |> 
  dplyr::rename("patient_n" = title,
                "gender" = characteristics_ch1,
                "disease_state" = characteristics_ch1.2,
                "identifier" = description.1)

meta_data |> 
  select(patient_n, gender, disease_state) |>
  sample_n(10)
```

We clean up patient, gender and disease state columns.

```{r}
meta_data <- meta_data |> 
  mutate(patient_n = str_extract(string = patient_n,
                                 pattern = "[0-9]+_?[0-9]?"),
         gender = str_extract(string = gender,
                              pattern = "Female|Male"),
         disease_state = str_extract(string = disease_state,
                              pattern = "HIV/TB|HIV"))
meta_data |> 
  sample_n(10)
```

The identifier is splited to simplify the identifier but still keeping the information about the dates.

```{r}
meta_data <- meta_data |> 
  separate(col = identifier,
           into = c("identifier", "date"),
           sep = "_")

meta_data |> 
  sample_n(10)
```

In the count data, the "-" in the identifier (column names) is a ".". We replace the "-" in the metadata:

```{r}
meta_data <- meta_data |> 
  mutate(identifier = str_replace_all(string = identifier,
                                      pattern = "-",
                                      replacement = "."))
```

We check if there is consensus between the identifiers in the meta data and the column names of the count data.

```{r}
#| label: Compare identifiers for metadata and count data 
meta_identifier <- meta_data |>
  pull(identifier)

setdiff(colnames(count_data),meta_identifier)
```

The only difference between identifiers and column names of the count data is the ID column for the genes as expected.

## Data Wrangling II

In the following steps, the format of the data will be changed. Furthermore, we wish to join the count data and the meta data.

### Data Pivoting

The count data is transposed by pivoting sample identifiers to long format and gene ids to wide format.

```{r}
count_data <- count_data |> 
  pivot_longer(cols = AAA0VKQN.12:EAA0ZNH3.13,
               names_to = "identifier",
               values_to = "count") |> 
  pivot_wider(names_from = ID,
              values_from = count)
```

### Data Joining

The meta data and the count data should be merge such that all data about a patient is contained in one tibble. Meta data should be in the first columns followed by gene variables.

```{r}
dat_clean <- inner_join(x = count_data,
                        y = meta_data,
                        by = "identifier") |> 
  relocate(colnames(meta_data))
```

### Handling of Replicates

There are two technical replicates available for one patient, namely patient 2_1 and 2_2 indicated by the patient_n column. We wish to averaged the measurements for this patient for further analysis. First, the gene expression values for the replicates are extracted.

```{r}
#| label: Extract replicates
dat_rep <- dat_clean |>
  filter(str_detect(patient_n, "_")) |>
  mutate(patient_n = str_extract(string = patient_n, 
                                 pattern = "[0-9]+")) 

dat_rep 
```

The mean gene expression is calculated for each replicated patient.

```{r}
#| label: Mean expression for replicates
#Calculate the mean expression across replicates of the same patient
dat_rep_mean <- dat_rep |> 
  group_by(patient_n) |>
  summarise(across(starts_with("ILMN"), 
                   mean, na.rm = TRUE)) #na.rm to exclude NA from calculations

dat_rep_mean

```

The metadata for each replicated patient is extracted. The "geo accession" and "identifer" is collapsed, meaning that the new "geo accession" and "identifer" for the replicated patient is a combination of the originals.

Furthermore, a new column called "modified" is added to the metadata. The column describes if the patient has been modified (eg. the gene expression represent a mean of multiple gene expressions for one patient).

```{r}
#| label: Extraxt and collapse metadata for replicates
cols_rep_meta_collapesed <- dat_rep |>
  group_by(patient_n) |> 
  summarise(geo_accession = paste0(geo_accession, collapse = "_"),
            identifier = paste0(identifier, collapse = "_"))

cols_rep_meta <- dat_rep |>
  select(patient_n, gender, disease_state, date) |> 
  distinct() |> 
  add_column(cols_rep_meta_collapesed |> select(geo_accession, identifier),
             modified = 1)
```

Combine the new metadata with the mean gene expression for the each replicated patient.

```{r}
#| label: Combine mean expression and metadata for replicates
dat_rep_meta <- inner_join(x = cols_rep_meta, 
                          y = dat_rep_mean,
                          by = "patient_n")
```

Finally, the new data for th replicated patients are combines with the data for the non-replicated patients to form a clean data set without replicates.

As the patient_n variable is no longer informative, this column is removed for all patients.

```{r}
#| label: Add rep data with non-rep data
dat_clean <- bind_rows(dat_clean, dat_rep_meta) |>
  filter(!str_detect(patient_n, "_")) |>  
  dplyr::select(-patient_n) |> 
  arrange(disease_state)

dat_clean |> 
  sample_n(10)
```

## Write Data to File

The cleaned data is written as a tsv file.

```{r}
dat_clean |> 
  write_tsv(file = "../data/02_dat_clean.tsv")
```
