---
title: "Describe data"
format: html
editor: visual
---

## Setup

Loading required packages

```{r}
#| label: Load packages
#| message: false
library("tidyverse")
library("table1")
```

Loading the cleaned data.

```{r}
#| label: Load data
#| message: false
dat_clean <- read_tsv("../data/02_dat_clean.tsv")
```

## Data Description

Create a bar plot to illustrate the number of males and females for each patient group.

```{r}
#| label: Define settings for plot
female_sign = intToUtf8(9792)
male_sign = intToUtf8(9794)
sign_size = 10
bar_width = 0.75
```

```{r}
#| label: Bar plot
groups <- dat_clean |> 
  group_by(gender,
           disease_state) |> 
  summarize(n = n())

barplot_patient_group <- 
  ggplot(data = dat_clean,
        mapping = aes(x = disease_state,
                      fill = disease_state,
                      alpha = gender)) +
  geom_bar(width = bar_width,
           position = "dodge") +
  scale_fill_manual(
    values = c("HIV" = "royalblue", 
               "HIV/TB" = "tomato") ) +
  scale_alpha_manual(values = c("Male" = 0.6,
                                "Female" = 0.9)) +
  annotate("text", 
           x = 1:2-bar_width/4, 
           y = groups$n[1:2] + 2, 
           label = female_sign, 
           size = sign_size) +
  annotate("text", 
           x = 1:2+bar_width/4,
           y = groups$n[3:4] + 2, 
           label = male_sign, 
           size = sign_size) +
  theme_minimal() + 
  labs(
    x = "Disease state",
    y = "Cases",
    fill = "Gender",
    title = str_c("Distribution of males and females for HIV mono-infected and HIV/TB",
                  "\n",
                  "co-infected patients"),
    caption = "Data from Dawany et al. (2014)"
  ) +
  theme(legend.position = "none")

ggsave("../results/barplot_patient_group.png", plot = barplot_patient_group)
show(barplot_patient_group)
```

The distribution of males and females for each patient group can also be summarized in a table.

```{r}
#| label: Summary table
summary_table <- dat_clean |>
  mutate(Gender = factor(gender),
         Disease = factor(disease_state)) |> 
  table1(x = formula(~ Gender | Disease))
show(summary_table)
```
From the above it is seen, that the gender of the HIV/TB co-infected patients is close to equal distributed between male and female. and that the gender of the HIV mono-infected is equal distributed with 72% females and only 27% males. 
